---
title: "Bayesian modeling of COVID19: Single Gompertz equation"
output: html_notebook
---
# * Version 0.2
# History
* Date: 2020.06.20 
* : Now Dynamical database from package covid19  (see https://covid19datahub.io)
* : package dplyr 

* Date 2020.06.28 
*  : functions for the MCMC
 
* Date 2020.06.29
* : Functions for plotting (R base) 
* : Sources file called helpers.R where the functions are.

* Date 2020.07.01
* : a loop to inform the priors by providing a MLE

* Date 2020.07.15
* Updated : Function rum_mcmc_simulation. Now uses only the country_index.
*  : For-loop on countries replaced by foreach %dopar%

# Helper functions
```{r}
source("./covid19_gompertz_utilities.R")
```

```{r}
# doFuture for countries
doFuture::registerDoFuture()
future::plan(multisession)
```

# Data cutoffs, directories, and other parameters : For more info, see `covid19_gompertz_utilities.R`
```{r}
confirmed_cases_cutoff         <- 1000              # Filters out countries with less confirmed cases than
fatalities_cutoff              <- 25                # Filters out countries with less fatalities than
density_confirmed_cases_cutoff <- 0.001             # Filters out countries with less density of confirmed cases than
population_cutoff              <- 1e6               # Filters out countries with less population than

# Directories and file names
results_dir      <- "/Users/ado/Desktop/COVID19_PROJECT/Results/Devs/"   
csv_file_fat_ind <- paste0(results_dir, "Summaries/", "fatalities_", date_cutoff,".csv")
csv_file_cc_ind  <- paste0(results_dir, "Summaries/", "confirmed_cases_", date_cutoff,".csv")

# MCMC variables
niter_sim_max  <- 1e4      # Exits with (geq). Recommended: 1e7

# Plotting 
plot_flag      <- 0        # 0 : no output on disk but on terminal. 
                           #     This flag also rules the writing of summaries.
```

# Data 
Data is loaded with one week (7 days) lag.
```{r}
full_data <- covid19(verbose = FALSE, level = 1, end = date_last)
```

Select columns of interest, rename them, and calculate the number of days elapsed from day_zero = Date[1]
```{r}
dat <-full_data %>% 
  select(id, Country = administrative_area_level_1, Date = date, Confirmed_Cases = confirmed, Fatalities = deaths, Population = population) %>% 
  group_by(Country) %>%
  mutate(Density_Confirmed_Cases = max(Confirmed_Cases) / Population, Time = Date - Date[1]) %>% 
  ungroup() 

day_zero <- dat$Date[1]
```

# Working dataframe
## Selecting countries using data cutoffs
Non-zero cut-offs for the number of `fatalities`, the `Confirmed_Cases` and the `density of confirmed cases`, and/or the `population` are important to define, because they allow for filtering of countries where the covid19 has not yet achieved the level of an epidemic. That is, countries that are in the initial stages of the disease or where it has been controlled at the early stage.  

For such countries there are no well defined trends, i.e., not enough information, to fit a Gompertz model for the evolution of confirmed cases and/or fatalities.
Filter out the countries by chosen cutoff and make sure the list of countries is current:

Select countries by a cutoff criterion (in this case "Pop+Dens"):
```{r}
(by_country<-include_countries(dat, "Pop+Dens"))
length(by_country)
```

## Selecting countries by name
Alternatively, a list of (at least two) countries can be provided by name, i.e., `by_country = c("Country1", "Country2", ...)`
```{r}
by_country <- c("France","Germany", "Romania", "Dominican Republic")
```

Re-filter the countries in `dat_by_country` to keep only the countries above the chosen cutoff:
```{r}
dat_by_country <- dat %>% 
  filter(Country %in% by_country) %>% 
  group_by(Country)  %>% 
  mutate(Daily_Confirmed_Cases = Confirmed_Cases - lag(Confirmed_Cases), Daily_Fatalities = Fatalities - lag(Fatalities)) %>%
  ungroup()
```

The number of days in each country under study is:
```{r}
nrow(dat_by_country)/ length(by_country)
```

Thus a variable `date_cutoff` is defined (see above in Section "Data cutoffs") as to separate the traning from the validation sets:
```{r}
train_by_country <- dat_by_country %>% filter(Date <= date_cutoff)
test_by_country  <- dat_by_country %>% filter(Date >  date_cutoff)
```

## Excluded countries
Number of countries off the list due to the `density_confirmed_cases_cutoff` and `population_cutoff` cutoffs:
```{r}
# Type = "Fatalities", "Confirmed_Cases", "Pop+Dens"
list_excluded_countries <- exclude_countries(dat,"Pop+Dens")
```

Create a dataframe with the excluded countries:
```{r}
dat_excluded_countries <- dat %>% 
  filter(Country %in% list_excluded_countries ) %>% 
  arrange(desc(Fatalities))

```

Arrange by the excluded countries by the maximum of confirmed cases:
```{r}
dat_excluded_countries %>% 
  filter(Date == date_last) %>% 
  group_by(Country) %>%
  arrange(desc(Confirmed_Cases))
```

Arrange by the excluded countries by the maximum of the density of confirmed cases:
```{r}
dat_excluded_countries %>% 
  filter(Date == date_last) %>% 
  group_by(Country) %>%
  arrange(desc(Density_Confirmed_Cases))
```


# Modeling
## Non-linear fit independent model per country
### Fatalities as function of time
```{r}
# Flat but sensible priors
mean_logasymptote_fatalities <- 1e1
mean_location_fatalities     <- 1e2
mean_growth_rate_fatalities  <- 0.01

prec_logasymptote_fatalities <- 1e-4	
prec_location_fatalities     <- 1e-4
prec_growth_rate_fatalities  <- 1e-4	

# Loop for the priors
for (index_prior_loop in 1:2){
  
# Fatalities model string
gompertz_fatalities_string <- build_model_string("Fatalities", 
                                                 mean_logasymptote_fatalities, 
                                                 mean_location_fatalities, 
                                                 mean_growth_rate_fatalities, 
                                                 prec_logasymptote_fatalities, 
                                                 prec_location_fatalities, 
                                                 prec_growth_rate_fatalities
                                                 )

# Parameters to trace
gompertz_fatalities_params <- c("logasymptote","location","growth_rate","prec")

# Begin MCMC simulation 
# Uses lowered values for the number_steps_* variables
registerDoRNG(1970)
fatalities_output_list <- foreach (i = 1:length(by_country)) %dopar% 
run_mcmc_simulation( index_country         = i , 
                     model_string          = gompertz_fatalities_string , 
                     model_parameters      = gompertz_fatalities_params,
                     number_steps_update   = 1e3, 
                     number_steps_run_init = 1e3, 
                     number_steps_run_max  = 1e3
                     )

# Create table of summaries 
fatalities_summary <- create_dummy_summary_df() %>% rename(Max_Fatalities = FATorConfCases)

# Summary countries loop
for (i in 1:length(by_country)){
  
# Fill-in the summary dataframe by row = country
gompertz_fatalities_csim <- fatalities_output_list[[i]][[2]]
final_niter_fatalities   <- fatalities_output_list[[i]][[3]]

fatalities_summary[i, ] <- summary_by_country("Fatalities", 
                                              gompertz_fatalities_csim, 
                                              by_country[i], 
                                              date_pred, 
                                              train_by_country, 
                                              test_by_country, 
                                              final_niter_fatalities)
} # End summary countries  loop

# Update the priors
fatalities_means <- fatalities_summary %>% 
  select(Growth_Rate, Location, Log_Asymptote, Precision)  %>%
  summarize(mean(Log_Asymptote), mean(Location), mean(Growth_Rate), mean(Precision))

fatalities_prec <- fatalities_summary %>% 
  select(Growth_Rate, Location, Log_Asymptote, Precision)  %>%
  summarize(1/var(Log_Asymptote), 1/var(Location),1/var(Growth_Rate), var(Precision))

mean_logasymptote_fatalities <- fatalities_means[[1]]
mean_location_fatalities     <- fatalities_means[[2]]
mean_growth_rate_fatalities  <- fatalities_means[[3]]

prec_logasymptote_fatalities <- fatalities_prec[[1]]	
prec_location_fatalities     <- fatalities_prec[[2]]
prec_growth_rate_fatalities  <- fatalities_prec[[3]]

 } # end loop priors

# Fatalities model string using the pre-converged priors
gompertz_fatalities_string <- build_model_string("Fatalities", 
                                                 mean_logasymptote_fatalities, 
                                                 mean_location_fatalities, 
                                                 mean_growth_rate_fatalities, 
                                                 prec_logasymptote_fatalities, 
                                                 prec_location_fatalities, 
                                                 prec_growth_rate_fatalities)

# Parameters to trace
gompertz_fatalities_params <- c("logasymptote","location","growth_rate","prec")
  
# Begin MCMC simulation 
# Notice that the number_steps_* variables are now the ones defined for the actual calculation (see preamble)
registerDoRNG(1970)
output_mcmc_fatalities <- foreach (i = 1:length(by_country)) %dopar% 
  run_mcmc_simulation( index_country    = i ,
                       model_string     = gompertz_fatalities_string , 
                       model_parameters = gompertz_fatalities_params 
                       )

# Create table of summaries 
fatalities_summary <- create_dummy_summary_df() %>% rename(Max_Fatalities = FATorConfCases)

# Summary countries loop
for (i in 1:length(by_country)){
  
# Fill the summary dataframe by row = country
gompertz_fatalities_csim  <- output_mcmc_fatalities[[i]][[2]]
final_niter_fatalities    <- output_mcmc_fatalities[[i]][[3]]

fatalities_summary[i, ] <- summary_by_country("Fatalities", 
                                              gompertz_fatalities_csim, 
                                              by_country[i], 
                                              date_pred, 
                                              train_by_country, 
                                              test_by_country, 
                                              final_niter_fatalities)
# Plot
plot_gompertz("Fatalities", train_by_country, Fatalities~Time, gompertz_fatalities_csim, projected_days, plot_flag = plot_flag)
plot_gompertz_daily_cases("Fatalities", train_by_country, Daily_Fatalities~Time, gompertz_fatalities_csim, projected_days, plot_flag = plot_flag)

} # End loop


# Save the results from tables / data frames
if (plot_flag == 1){
write.csv(fatalities_summary, csv_file_fat_ind)
}

# A summary of the Gompertz parameters for the independent models
# fatalities_summary
```

### The number of Confirmed Cases as function of time
```{r}
# Flat but sensible priors
mean_logasymptote_confirmed_cases <- 1e1
mean_location_confirmed_cases     <- 1e2
mean_growth_rate_confirmed_cases  <- 0.01

prec_logasymptote_confirmed_cases <- 1e-4	
prec_location_confirmed_cases     <- 1e-4
prec_growth_rate_confirmed_cases  <- 1e-4	

# Loop for the priors
for (index_prior_loop in 1:2){
  
# Confirmed_Cases model string
gompertz_confirmed_cases_string <- build_model_string("Confirmed_Cases", 
                                                 mean_logasymptote_confirmed_cases , 
                                                 mean_location_confirmed_cases , 
                                                 mean_growth_rate_confirmed_cases , 
                                                 prec_logasymptote_confirmed_cases , 
                                                 prec_location_confirmed_cases , 
                                                 prec_growth_rate_confirmed_cases )

# Parameters to trace
gompertz_confirmed_cases_params <- c("logasymptote","location","growth_rate","prec")
  
# Begin MCMC simulation 
# Uses lowered values for the number_steps_* variables
registerDoRNG(1970)
confirmed_cases_output_list<- foreach (i = 1:length(by_country)) %dopar% 
  run_mcmc_simulation( index_country         = i , 
                       model_string          = gompertz_confirmed_cases_string , 
                       model_parameters      = gompertz_confirmed_cases_params,
                       number_steps_update   = 1e3, 
                       number_steps_run_init = 1e3, 
                       number_steps_run_max  = 1e3
                       )

# Create table of summaries 
confirmed_cases_summary <- create_dummy_summary_df() %>% rename(Max_Confirmed_Cases = FATorConfCases)

# Summary countries loop
for (i in 1:length(by_country)){
    
# Fill-in the summary dataframe by row = country
gompertz_confirmed_cases_csim <- confirmed_cases_output_list[[i]][[2]]
final_niter_confirmed_cases   <- confirmed_cases_output_list[[i]][[3]]
    
confirmed_cases_summary[i, ] <- summary_by_country("Confirmed_Cases", 
                                                   gompertz_confirmed_cases_csim, 
                                                   by_country[i], 
                                                   date_pred, 
                                                   train_by_country, 
                                                   test_by_country, 
                                                   final_niter_confirmed_cases)
                                              
} # End summary countries loop

# Update the priors
confirmed_cases_means <- confirmed_cases_summary %>% 
  select(Growth_Rate, Location, Log_Asymptote, Precision)  %>%
  summarize(mean(Log_Asymptote), mean(Location), mean(Growth_Rate), mean(Precision))

confirmed_cases_prec <- confirmed_cases_summary %>% 
  select(Growth_Rate, Location, Log_Asymptote, Precision)  %>%
  summarize(1/var(Log_Asymptote), 1/var(Location),1/var(Growth_Rate), var(Precision))

mean_logasymptote_confirmed_cases <- confirmed_cases_means[[1]]
mean_location_confirmed_cases     <- confirmed_cases_means[[2]]
mean_growth_rate_confirmed_cases  <- confirmed_cases_means[[3]]

prec_logasymptote_confirmed_cases <- confirmed_cases_prec[[1]]	
prec_location_confirmed_cases     <- confirmed_cases_prec[[2]]
prec_growth_rate_confirmed_cases  <- confirmed_cases_prec[[3]]

} # end loop priors

# Confirmed_Cases model string using the pre-converged priors
gompertz_confirmed_cases_string <- build_model_string("Confirmed_Cases", 
                                                 mean_logasymptote_confirmed_cases , 
                                                 mean_location_confirmed_cases , 
                                                 mean_growth_rate_confirmed_cases , 
                                                 prec_logasymptote_confirmed_cases , 
                                                 prec_location_confirmed_cases , 
                                                 prec_growth_rate_confirmed_cases )

# Parameters to trace
gompertz_confirmed_cases_params <- c("logasymptote", "location", "growth_rate", "prec")

# Begin MCMC simulation 
# Notice that the number_steps_* variables are now the ones defined for the actual calculation (see preamble)
registerDoRNG(1970)
output_mcmc_confirmed_cases <- foreach (i = 1:length(by_country)) %dopar% 
  run_mcmc_simulation( index_country    = i , 
                       model_string     = gompertz_confirmed_cases_string , 
                       model_parameters = gompertz_confirmed_cases_params 
                       )

# Create table of summaries 
confirmed_cases_summary <- create_dummy_summary_df() %>% rename(Max_Confirmed_Cases = FATorConfCases)

# Summary countries loop
for (i in 1:length(by_country)){
  
# Fill the summary dataframe by row = country
gompertz_confirmed_cases_csim <- output_mcmc_confirmed_cases[[i]][[2]]
final_niter_confirmed_cases   <- output_mcmc_confirmed_cases[[i]][[3]]

confirmed_cases_summary[i, ] <- summary_by_country("Confirmed_Cases", 
                                                   gompertz_confirmed_cases_csim, 
                                                   by_country[i], 
                                                   date_pred, 
                                                   train_by_country, 
                                                   test_by_country, 
                                                   final_niter_confirmed_cases)

# Plot
plot_gompertz("Confirmed_Cases", train_by_country, Confirmed_Cases~Time, gompertz_confirmed_cases_csim, projected_days, plot_flag = plot_flag)
plot_gompertz_daily_cases("Confirmed_Cases", train_by_country, Daily_Confirmed_Cases~Time, gompertz_confirmed_cases_csim, projected_days, plot_flag = plot_flag)

} # End countries loop

## Save the results from tables / data frames
if (plot_flag == 1){
write.csv(confirmed_cases_summary, csv_file_cc_ind)
}

# A summary of the Gompertz parameters for the independent models
# confirmed_cases_summary
```

