---
title: "Bayesian modeling of COVID19: Double Gompertz equation"
output: html_notebook
---
# * Version 1.1
Uses data from package covid19  (see https://covid19datahub.io) to fit single and double Gompertz equations to model one- and two-wave  COVID1-9 spreading, respectively, by means of a Bayesian MCMC approach.

# Preliminaries
```{r}
# Helper functions and data
source("./covid19_gompertz_utilities.R")
source("./covid19_gompertz_parameters.R")
# Path to save results
results_path <- file.path("" , "dir1" , "dir2" , "dir3" )
```

# Data from covid19datahub
* `full_data`             : Loaded with one week (2 days) lag. 
* `sf_dat`                : `full_data`-selected and fortified.
* `latest_global_figures` : Latest figures worldwide from `sf_dat`.
* `dat`                   : Working database `sf_dat`-country-filtered by Pop+Dens cutoffs (see below)

Database `latest_global_figures`
```{r}
latest_global_figures
```

Database `dat` has been filtered by following thresholds
```{r}
print(paste("Confirmed cases >"           , confirmed_cases_cutoff))
print(paste("Density of confirmed cases >", density_confirmed_cases_cutoff))
print(paste("Country's population >"      , population_cutoff))
```

A list of the countries included in the working database `dat` can be obtained using `include_countries(sf_dat, "Pop+Dens")` and for those excluded `exclude_countries(sf_dat, "Pop+Dens")`. Also useful to check the spelling and capitalization of countries names.

Latest (per `date_last`) top-10/bottom-10 country-ranking is possible for the following (13) variables
```{r}
latest_figures <- get_latest_figures(dat) 
names(latest_figures)[-c(1:3,17)]
```

For instance, `Fatalities`, `Density_Fatalities`, and `Daily_Fatalities`
```{r}
top_10_ranking(df = latest_figures, type = "Recovered")
top_10_ranking(df = latest_figures, type = "Daily_Fatalities")
top_10_ranking(df = latest_figures, type = "Density_Confirmed_Cases")
```

Or all of them via the function `screen_top_10()`
```{r}
screen_top_10(df = latest_figures)
```

Storing the plots using `store_top_10()`
```{r}
store_top_10(df =latest_figures , save_path = results_path, res = 180)
```

Countries sorted by wave of `Confirmed_Cases`, `Fatalities`, and `Recovered` :
```{r}
list_of_countries <- include_countries(dat, "Pop+Dens")
countries_by_wave <- rank_countries_by_wave(list_of_countries, dat)
```

Countries of interest can also be selected individually :
```{r}
by_country <- c("France", "Mexico", "Romania")
```

# MCMC simulation
MCMC estimation of a Gompertz function for each of the `type` in `by_cases` and countries in `by_country`. Active cases are not collected directly by the covid19datahub. However, the number of active cases can by estimated as the difference between the confirmed cases and the sum of fatalities and recovered. 

```{r}
# Needed to execute the MCMC and visualization
fit_range <- 0:projected_days

# Initialize list of types/cases 
by_cases              <- c("Confirmed_Cases", "Fatalities", "Recovered")
case_list_output_mcmc <- vector(mode = "list", length = length(by_cases))

for (by_type in by_cases){
  
  # Initialize 
  country_list_output_mcmc <- vector(mode = "list", length = length(by_country))

  # Flat but sensible priors
  mean_logasymptote_1 <- 1e1
  mean_location_1     <- 1e2
  mean_growth_rate_1  <- 0.01
  prec_logasymptote_1 <- 1e-4  
  prec_location_1     <- 1e-4
  prec_growth_rate_1  <- 1e-4

  mean_logasymptote_2 <- 1e1
  mean_location_2     <- 1e2
  mean_growth_rate_2  <- 0.01
  prec_logasymptote_2 <- 1e-4  
  prec_location_2     <- 1e-4
  prec_growth_rate_2  <- 1e-4


  registerDoRNG(1970)
  country_list_output_mcmc <- foreach (i = 1:length(by_country)) %dopar% {
  
    by_wave <- get_wave( df      = countries_by_wave , 
                         country = by_country[i]     , 
                         type    = by_type           , 
                         fswave  = 1
                         ) 
    
    # This how to override the automatic wave classification:
    if (by_country[i] == "Canada" & by_type == "Fatalities"){by_wave <- 1}
    if (by_country[i] == "Mexico"){by_wave <- 1}
    
    list_gompertz_string <- list( type                = by_type             , 
                                      
                                  mean_logasymptote_1 = mean_logasymptote_1 ,
                                  mean_location_1     = mean_location_1     , 
                                  mean_growth_rate_1  = mean_growth_rate_1  , 
                              
                                  prec_logasymptote_1 = prec_logasymptote_1 , 
                                  prec_location_1     = prec_location_1     , 
                                  prec_growth_rate_1  = prec_growth_rate_1  ,
                                     
                                  mean_logasymptote_2 = mean_logasymptote_2 ,
                                  mean_location_2     = mean_location_2     , 
                                  mean_growth_rate_2  = mean_growth_rate_2  , 
                              
                                  prec_logasymptote_2 = prec_logasymptote_2 , 
                                  prec_location_2     = prec_location_2     , 
                                  prec_growth_rate_2  = prec_growth_rate_2  ,
                                      
                                  wave                = by_wave
                                  )
  
  
    run_mcmc_simulation_2( index_country         = i                    , 
                           list_model_string     = list_gompertz_string , 
                           number_chains         = 1                    ,
                           number_steps_update   = 1e4                  , 
                           number_steps_run_init = 2e3                  , # 2-5e3
                           number_steps_run_max  = 1e6                    # 1e6
                           )
  }

  names(country_list_output_mcmc)                   <- by_country
  case_list_output_mcmc[[match(by_type, by_cases)]] <- country_list_output_mcmc
}

names(case_list_output_mcmc) <- by_cases
remove(country_list_output_mcmc)
gc()

gfit_active_cases <- vector(mode = "list", length = length(by_country))

gfit_active_cases <- foreach (i = 1:length(by_country)) %dopar% { 
  gfit <- gompertz_eval( case_list_output_mcmc[["Confirmed_Cases"]][[by_country[i]]][["csim"]] , 
                 fit_range                                                             ,
                 case_list_output_mcmc[["Confirmed_Cases"]][[by_country[i]]][["wave"]]
                 ) -
  gompertz_eval( case_list_output_mcmc[["Fatalities"]][[by_country[i]]][["csim"]] ,
                 fit_range                                                        ,
                 case_list_output_mcmc[["Fatalities"]][[by_country[i]]][["wave"]] 
                 ) -
  gompertz_eval( case_list_output_mcmc[["Recovered"]][[by_country[i]]][["csim"]] ,
                 fit_range                                                       ,
                 case_list_output_mcmc[["Recovered"]][[by_country[i]]][["wave"]] 
                 ) 
  return_list <- list(csim = gfit, wave = 1) 
  }

names(gfit_active_cases)     <- by_country
case_list_output_mcmc[[4]]   <- gfit_active_cases
by_cases                     <- c(by_cases, "Active_Cases")
names(case_list_output_mcmc) <- c(by_cases)

remove(gfit_active_cases)
```

# Visualization and postprocessing
```{r}
# Initialize 
case_list_output_vis <- vector(mode = "list", length = length(by_cases))

for (by_type in by_cases){
  country_list_output_vis <- vector(mode = "list", length = length(by_country))

  country_list_output_vis <- foreach (i = 1:length(by_country)) %dopar% {

    visualize_cases_mcmc( country_index    = i                      ,  
                          type             = by_type                , 
                          df               = dat                    , 
                          x                = fit_range              ,
                          list_output_mcmc = case_list_output_mcmc[[by_type]][[by_country[i]]] , 
                          wave             = case_list_output_mcmc[[by_type]][[by_country[i]]] 
                          )
  }
  
  names(country_list_output_vis) <- by_country
  case_list_output_vis[[match(by_type, by_cases)]] <- country_list_output_vis
}

names(case_list_output_vis) <- by_cases
remove(country_list_output_vis)
gc()
```

Visualizing the results by calling directly each element of the list `case_list_output_vis` (notice that currently you can choose `"Case"`, `"Case_Zoom"`, `"Daily_Case"` and `"Daily_Case_Zoom"`):
```{r}
case_list_output_vis[["Fatalities"]][[by_country[1]]][["Case"]]
case_list_output_vis[["Fatalities"]][[by_country[1]]][["Case_Zoom"]]
```

By country:
```{r}
invisible( lapply( by_cases, function(type) print(case_list_output_vis[[type]][[by_country[1]]][["Case"]])       ))
invisible( lapply( by_cases, function(type) print(case_list_output_vis[[type]][[by_country[1]]][["Daily_Case"]]) ))
```

All of them via the function `screen_plots()`
```{r}
screen_plots(case_list_output_vis)
```

Saving the results to disk :
```{r}
store_plots( case_list_output_vis,
             save_path = results_path,
             res       = 180
             )
```

